A Few Easy Pieces: \\ How to enhance your relationship with your future self.

### 2011-02-27 workflow.Rnw
Added vc files. A few more edits. Still not a first draft.

\title{A paper is to write}

\author{Jake Bowers\thanks{Many thanks to Kevin Quinn and Cara Wong for direct help on this document, Mika LaVaque-Manty, Ben Hansen, and Mark Fredrickson for many useful discussions on this topic, and Kieran Healy for inspiration from afar.}}

@@

options(SweaveHooks=list(
          fig=function(){
            par(mar=c(3.5, 3, 1.1, 0),
                pty="s",
                mgp=c(1.5,0.5,0),
                oma=c(0,0,0,0))},#,
          echo=function(){options(continue=" ") ##Don't show "+" prompts,
                         options(prompt=" ")}
          ),
        digits=3,
        scipen=8,
        width=100
        )

@@

The first author convenes a three way video conference with the other collaborators while commuting home in his flying car. The group must go back to the analyses. Which ones? The ones from 2011? Or 2012? Or 2018? Where are the files? The next day, one member of the group finds some of the files. Now the checks and re-analyses should be easy. Right? The student-now-faculty member is the same person and should remember the reason for those bits of code (or at least should remember which series of mouse clicks were used to produce the numbers for that crucial table). And, of course, the way Microsoft Word/Stata/SPSS/R/LISREL stores files and the way that your Windows/Mac machine reads and writes them is the same, right? (And Windows and Mac are still relevant operating systems.) And the group knows exactly which bit of code produced which table and which figure, right? And they wrote their code following Nagler's Maxims \citet{nagler1995coding} and being following King's Replication Standard \citet{king1995replication} right?

@@

\section{Why should we believe each other?}

When we call some piece of work ``scientific'' we are talking about how it was done. By calling it ``scientific'' we are also using a short-hand to suggest that (1) the findings of the work are not a matter of opinion and (2) other people could reproduce the findings. That is, what the work represents is not a matter of opinion but is a shared experience --- and an experience shared without respect to the identities of others (although requiring some shared technical training and research resources). This short-hand means that ``scientific'' work should make us change how we act --- or at least, it ought to stand on stronger epistemological ground than other claims about experiences which ought to be shared or shareable.

Assume we want other scientists to believe us when we say something. More narrowly, assume we want other people to believe us when we say something about data: ``data'' here can be words, numbers, musical notes, images, ideas, etc \ldots The point is that we are making some claims about patterns that we see in some collection of stuff. Now, it might be easy to convince other people that ``this collection of stuff is different from this collection of stuff'' if those people were looking over our shoulders the whole time that we made decisions about collecting the stuff and analyzed it [where ``analyze'' means something like ``break it up into understandable parts'']. Unfortunately, we can't assume that people are willing to shadow a researcher throughout her career. Rather, we do our work alone or in small groups and want to convince other distant people about our analyses.

\section{Show me the data.}
All files containing commands operating on data must refer to a data file. A reference to a data file is a line of code the analysis program will use to operate on (``load''/ ``open'' / ``get'' / ``use'') the data file. One should not have to edit this line on different computers or platforms in order to execute this command.  Using R, for example, all analysis files should have load('thedata.rda') or \Verb+read.csv('http://www.mywebsite.org/Data/thedata.csv')+ or some equivalent line in them, and 'thedata.csv' should be stored in some place easy to find (like in the same directory as the file or perhaps in 'Data/thedata.rda').

@@

The principle regarding data is to know where the data came from and what operations were performed on which set of data. I have seen command files which do not begin with a call to load data and I have despaired. What is the meaning of \Verb+please-fit(var1 with var2)+ if I do not know where var1 var2 come from.\footnote{The command \Verb+please-fit+ comes from the MayIPleaseDoStatistics package for R which emphasizes politeness in data analysis. It is highly recommended for use with toddlers.}

\section{Code not Clicking}
All tables and graphs should be as much as possible generated from code and not clicking (let alone copying and pasting). If I wanted to re-create the figure you created but with red lines, and only for people in the South, I should be able to do so with just a few edits to the code.

@@

\begin{verbatim}
pdf('fig1.pdf')
please-plot(outcome by explanatory)
dev.off()
\end{verbatim}

Now, in the future if I wonder how ``that plot on page 10'' was created, if I suitably commented my manuscript or inserted text into my MANIFEST.txt file I will know (1) ``that plot'' is from a file called ``fig1.pdf'' and (2) fig1.pdf was created in \Verb+figs1and2.R+. In the future, if I wanted to quickly change fig1.pdf, I could edit the commands quickly and easily to do so. [Another good practice is to name the files descriptively, so that we don't in fact have ``fig1.pdf'' but ``political participation by education in the south.pdf'' although such a name may prove burdensome to operating systems which do not care for spaces in file names or to folks who are used to typing full filenames rather than using command line completion facilities in modern shells.]

@@

\begin{figure}[ht]
  \begin{center}
<<fig1plot,echo=FALSE,fig=TRUE,width=3,height=3>>= 
<<fig1code>>
@
    \caption{Protest activity by income inequality from \citep{norris2009data}.}\label{fig:giniprot}
  \end{center}
\end{figure}

@@

\section{Fishing rods and not fish}
The source code of any pdf document exchanged by the group must be available and executable. Say you run some command and then make a table or a graph. Then you produce a nice little report on your work for use among the working group. Eventually we want to use pieces of that report (tables, graphs, paragraphs) in a publishable paper. Thus, that report must not have been created using copying and pasting. It must be created using LaTeX (or HTML or OpenOffice or something else that the working group standardizes on). And, in fact, it must be created using the reproducible markup system like Sweave (if you are using LaTeX and R as your working group tools) \citep{Leis:2005}. Alternatively, you can put comments into your report detailing where the figures and tables came from (i.e. which command file produced them). Or include details about which code produced which figures in your MANIFEST.txt.

@@

The Sweave system is even nicer because you can have a LaTeX file with R code right inside of it! I showed Sweave with plotting above so here is an example in which a regression table is produced automagically:

<<maketable, echo=FALSE, results=tex>>=
##source("http://artsci.wustl.edu/~jgill/Models/graph.summary.R")

lm1<-lm(protac2000~I(Gini2004/100)+meanpr,data=good.df)

themat<-cbind(summary(lm1)$coef[,c("Estimate","Std. Error")],confint(lm1))
n<-length(lm1$fitted.values)
est.sigma<-round(summary(lm1)$sigma,3)
library(xtable)

colnames(themat)[3:4]<-c("~{}","~{}")
row.names(themat)<-c("Intercept","Income Inequality (lower=more equal)","Mean Political Rights (lower=more rights)")
thextab<-xtable(themat,align=c("l",rep(".",4)),
                caption="People living in countries with unequal
  income distributions report less protest activity to World Values
  Survey interviewers than people living in countries with relatively
  more equal income distributions, adjusting for average political
  rights as measured by Freedom House 1980--2000. Data from \\citep{norris2009data}.",label="tab:protest")

print(thextab,include.colnames=FALSE,
      hline.after=NULL,
      add.to.row=list(pos=list(0,3),
                command=c("\\toprule \n & \\multicolumn{1}{c}{Coef} & \\multicolumn{1}{c}{Std. Err.} 
& \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ", 
                  paste("& \\multicolumn{4}{c}{n: ",n,",   $\\hat{\\sigma}$:",est.sigma,"} \\\\ \\bottomrule"))
                )
      )
@

@@

\section{Enabling revision. Preventing clobbering.}

Group work requires version control.\footnote{See X for more
  discussion of what version control is.} Many people are familiar
with ``Track Changes'' feature in modern WYSIWYG word processors or the fact that
Dropbox allows one to recover previous versions of files. These are
both kinds of version control. When collaborating with yourself or
others, it is useful to see what has changed, to feel free to
experiment and then to dump the experiment in favor of previous work,
to have multiple ``releases'' of the same document (one to MPSA, one
to APSR, one to your parents) without requiring that spawn many copies
of the same document and risk confusion and clobbering. Clobbering is
what happens when your future self or your other collaborators saves
an old version of a file over a new version --- erasing good work by
accident.

@@

Fancy version control systems are not required, however. I suspect
that Google Docs allows a kind of version tracking and collaboration
as well.  An excellent, simple, and robust version control system is
to merely rename your files with the date and time of saving them:
thedoc.tex becomes thedoc25092009-22:40.tex.  If you are wise enough to have saved your documents as plain text (Such as LaTeX source (with or without R/Stata code chunks) \footnote{A quick Google search of ``Sweave for Stata'' turned up lots of resources for literate programming with Stata.} then you can easily compare documents using the many utilities available for comparing text files [FileMerge on OS X is pretty, but ediff and diff for Unix are very useful, I am sure that Windows has many other options]. Adobe Acrobat allows one to compare differences in pdf files. OpenOffice supports a ``Compare Documents'' option.

@@

\begin{Verbatim}
<<reg1>> =
##Fitting the regression with x/1000 for ease of interpretation
thelm<-lm(y~I(x/1000)+z,data=fabulous.df)
##Fits using a quadratic term as suggested by Blah(1865) did not enhance the fit of the model
##thelmsq<-lm(y~I(x/1000)^2+z,data=fabulous.df)
@
\end{Verbatim}

If you need to sleuth down something more than a small detail, but yet the sleuthing is really a footnote to the article, then you can create another file to explore and report on those questions for the research group. The writing in such a memo can be more informal, but the norms of reproducibility ought to be the same as for the main paper. After all you will be releasing both documents (probably) publicly upon publication of the article if not before.

@@

Name your files with evocative and descriptive names. Your
collaborators are less likely to call you at midnight asking for help
if your files are named ``regressions.R'' than if your files are
called ``temp9'' or ``supercalifragilisticexpialidocious'' (note the
use of the extension .R to tell us that the file contains R
commands. use extensions like this as a standard practice to help you
and your computer get along.)

(other stuff I've forgotten?)

@@

I have experimented with three systems so far: (1) for one paper we simply included a Sweave document and data files into a compressed archive \cite{bowers2005dataverse}; (2) for another more computing intensive paper we assembled a set of files which enabled reproduction of our results using the ``make'' system \citep{bowers2008dataverse}; and (3) recently I have tried the ``compendium'' approach \citep{gentleman2005reproducible,gentleman2007statistical} which embeds an academic paper with the R package system \cite{bowers2011dataverse}. The benefit of this last system is that one is not required to have access to unix: one may download, run, and view the paper from within R --- which is platform independent.

@@

We all always collaborate. Many of us collaborate with groups of people at one moment in time as we race to produce a paper in time for some deadline. All of us collaborate with ourselves over time. An idealized version of self-collaboration goes something like this: We begin a project with a few notes. We write a conference paper. We revise this paper and send it to a journal. The journal asks for revisions. We revise and resubmit. Our paper is accepted. We revise a bit more. The paper is published. Later another scholar reacts to our paper. We write another paper based on said reactions (draft, conference paper, submission, revision, acceptance, revision). These processes occur over years of time. So, I collaborate with a self 5 or 10 or even more years hence when I write a paper. [Ask anyone who has turned their dissertation into a book how long it took from them from dissertation prospectus to physical published book for an example of a common time span in which self-collaboration is required.]

The time-frames over which collaboration are required --- whether
among a group of people working together or within a single scholar's
productive life --- are much longer than any given version of any
given software will easily exist. All except for plain text. Thus,
even as we extol version control systems, one must have a way to
ensure future access to them in a form that will still be around when
the cockroaches take over political science departments.\footnote{
  (and thereby finally make Ctrl-a Ctrl-x Esc-x Ctrl-c a reasonable
  way to instruct Emacs to do something
  (\url{http://kieran.healy.usesthis.com/} )}

In the end, following these practices allows your analyses of your collections to be credible --- and, in the case of criticism, you can always tell antagonists to go redo your analysis if they really don't believe you.  Let me end with the most pithy statement in favor of the idea that collaboration is at the heart of our endeavor that I have found:

>>>>

\title{A Few Easy Pieces: \\ How to enhance your relationship with your future self.}

\author{Jake Bowers\thanks{Many thanks to Kevin Quinn and Cara Wong for direct help on this document, Mika LaVaque-Manty, Ben Hansen, and Mark Fredrickson for many useful discussions on this topic, and Kieran Healy for inspiration from afar. The source code for this document may be freely downloaded from \url{https://github.com/jwbowers/workflow}.}}

@@

options(SweaveHooks=list(
          fig=function(){
            par(mar=c(3.5, 3, 1.1, 0),
                pty="s",
                mgp=c(1.5,0.5,0),
                oma=c(0,0,0,0))},#,
          echo=function(){options(continue=" ") ##Don't show "+" prompts,
                         options(prompt=" ")}
          ),
        digits=3,
        scipen=8,
        width=100,
       error=function(){options(prompt="> ",continue="+ ");NULL}
        )

@@

The first author convenes a three way video conference with the other collaborators while commuting home in his flying car. The group must go back to the analyses. Which ones? The ones from 2011? Or 2012? Or 2018? Where are the files? The next day, one member of the group finds some of the files. Now the checks and re-analyses should be easy. Right? The student-now-faculty member is the same person and should remember the reason for those bits of code (or at least should remember which series of mouse clicks were used to produce the numbers for that crucial table --- and those mouse clicks should not be time consuming to redo exactly as they were done in 2011, or was it 2015?). And, of course, the way Microsoft Word/Stata/SPSS/R/LISREL stores files and the way that your machine reads and writes them is the same, right --- since Windows and Mac OS X have always existed and will always continue to exist more or less as they current exist. And the group knows exactly which bit of code produced which table and which figure, right? And they wrote their code following Nagler's Maxims \citet{nagler1995coding} and being following King's Replication Standard \citet{king1995replication} right?

@@

\section{Meaningful code requires data.}
All files containing commands operating on data must refer to a data file. A reference to a data file is a line of code the analysis program will use to operate on (``load''/ ``open'' / ``get'' / ``use'') the data file. One should not have to edit this line on different computers or platforms in order to execute this command.  Using R, for example, all analysis files should have load('thedata.rda') or \Verb+read.csv('http://www.mywebsite.org/Data/thedata.csv')+ or some equivalent line in them, and 'thedata.csv' should be stored in some place easy to find (like in the same directory as the file or perhaps in 'Data/thedata.rda').

@@

The principle regarding data is to know where the data came from and what operations were performed on which set of data. I have seen command files which do not begin with a call to load data and I have despaired. What is the meaning of \Verb+please-fit(var1 with var2)+ if I do not know where var1 var2 come from.\footnote{The command \Verb+please-fit+ and some other R functions used in this essay come from the MayIPleaseDoStatistics package for R which emphasizes politeness in data analysis. It is highly recommended for use with toddlers.}

\section{Type, don't click.}
All tables and graphs should be generated from code and not clicking (let alone copying and pasting). If I wanted to re-create the figure you created but with red lines, and only for people in the South, I should be able to do so with just a few edits to the code.

@@

\begin{Verbatim}[fontsize=\small]
pdf('fig1.pdf')
please-plot(outcome by explanatory)
dev.off()
\end{Verbatim}

Now, in the future if I wonder how ``that plot on page 10'' was created, if I suitably commented my manuscript or inserted text into my MANIFEST.txt file I will know (1) ``that plot'' is from a file called ``fig1.pdf'' and (2) fig1.pdf was created in \Verb+figs1and2.R+. In the future, if I wanted to quickly change fig1.pdf, I could edit the commands quickly and easily to do so. Another good practice is to name the files descriptively, so that we don't in fact have ``fig1.pdf'' but ``political participation by education in the south.pdf'' although such a name may prove burdensome to operating systems which do not care for spaces in file names or to folks who are used to typing full filenames rather than using command line completion facilities in modern shells.

@@

\begin{figure}[H]
  \begin{center}
<<fig1plot,echo=FALSE,fig=TRUE,width=3,height=3>>= 
<<fig1code>>
@
    \caption{Protest activity by income inequality \citep[from]{norris2009data}.}\label{fig:giniprot}
  \end{center}
\end{figure}

@@

\section{Enjoy literate data analysis.}
The source code of any pdf document exchanged by the group must be available and executable. Say you run some command and then make a table or a graph. Then you produce a nice little report on your work for use among the working group. Eventually we want to use pieces of that report (tables, graphs, paragraphs) in a publishable paper. Thus, that report must not have been created using copying and pasting. It must be created using LaTeX (or HTML or OpenOffice or something else that the working group standardizes on). And, in fact, it must be created using the reproducible markup system like Sweave (if you are using LaTeX and R as your working group tools) \citep{Leis:2005}. Alternatively, you can put comments into your report detailing where the figures and tables came from (i.e. which command file produced them). Or include details about which code produced which figures in your MANIFEST.txt.

@@

The Sweave system is even nicer because you can have a LaTeX file with R code right inside of it! I showed Sweave with plotting in Figure~\ref{fig:giniplot}, those who view the code for this essay will see how the Table~\ref{tab:protest} was also generated directly from a regression object.

<<maketable, echo=FALSE, results=hide>>=
lm1<-lm(protac2000~I(Gini2004/100)+meanpr,data=good.df)  #  Run the regression

##  The next lines extract relevant and useful output from the
##  regression object, following Neal Beck's TPM article 2010.
themat<-cbind(summary(lm1)$coef[,c("Estimate","Std. Error")],confint(lm1))
n<-length(lm1$fitted.values)
r2<-signif(summary(lm1)$r.squared,2)
est.sigma<-signif(summary(lm1)$sigma,2)
colnames(themat)[3:4]<-c("~{}","~{}")
row.names(themat)<-c("Intercept","Income Inequality (lower=more equal)","Mean Political Rights (lower=more rights)")

library(xtable)
thextab<-xtable(themat,align=c("l",rep(".",4)),
                caption="People living in countries with unequal
  income distributions report less protest activity to World Values
  Survey interviewers than people living in countries with relatively
  more equal income distributions, adjusting for average political
  rights as measured by Freedom House 1980--2000. Data from \\citep{norris2009data}.",label="tab:protest") 
@

<<printtable,echo=FALSE,results=tex>>= 
print(thextab,include.colnames=FALSE,
      hline.after=NULL,
      table.placement="H",
      add.to.row=list(
        pos=list(0,3),
        command=c(
                  "\\toprule \n &
 \\multicolumn{1}{c}{Coef} &
 \\multicolumn{1}{c}{Std. Err.} &
 \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ", 
                  paste("& \\multicolumn{4}{c}{n: ",n,",","   resid. sd:",signif(est.sigma),", R$^2$:",r2,"} \\\\ \\bottomrule",sep="")
                  )
        ))
@

@@

\section{Enabling revision to prevent clobbering.}

Group work requires version control.\footnote{See X for more
  discussion of what version control is.} Many people are familiar
with the ``Track Changes'' feature in modern WYSIWYG word processors or the fact that
Dropbox allows one to recover previous versions of files. These are
both kinds of version control. When collaborating with yourself or
others, it is useful to see what has changed, to feel free to
experiment and then to dump the experiment in favor of previous work,
to have multiple ``releases'' of the same document (one to MPSA, one
to APSR, one to your parents) without requiring that spawn many copies
of the same document and risk confusion and clobbering. Clobbering is
what happens when your future self or your other collaborators saves
an old version of a file over a new version --- erasing good work by
accident.

@@

Fancy version control systems are not required, however. I suspect
that Google Docs allows a kind of version tracking and collaboration
as well.  An excellent, simple, and robust version control system is
to merely rename your files with the date and time of saving them:
thedoc.tex becomes thedoc25-12-2011-23:50.tex.  Be sure to include
year in the file names --- remember, the life of an idea is measured
in years. Also, if you use this method, spend a little extra time to
ensure that you do not clobber files when you make typos in the
file-names.

If you are wise enough to have saved your documents as plain text (Such as LaTeX source (with or without R/Stata code chunks) \footnote{A quick Google search of ``Sweave for Stata'' turned up lots of resources for literate programming with Stata.} then you can easily compare documents using the many utilities available for comparing text files [FileMerge on OS X is pretty, but ediff and diff for Unix are very useful, I am sure that Windows has many other options]. Adobe Acrobat allows one to compare differences in pdf files. OpenOffice supports a ``Compare Documents'' option.

@@

<<reg1,echo=TRUE,results=verb>>=
##  Repeat the regression run for Table~\ref{tab:protest} [use a label instead
##  of a number to make moving sections and tables easier]
##  Do political rights predict protest reporting?
lm1<-lm(protac2000 ~ I(Gini2004/100) + meanpr, good.df)
##  Q: Was it worthwhile controlling for political rights?
lm2<-lm(protac2000~I(Gini2004/100),data=good.df)
coef(lm2)[2]-coef(lm1)[2]
##  A: Yes (around 3/4 an act of protest difference)
@

If you need to track down something more than a small detail, but yet the sleuthing is really a footnote to the article, then you can create another file to explore and report on those questions for the research group. The writing in such a memo can be more informal, but the norms of reproducibility ought to be the same as for the main paper. After all you will be releasing both documents (probably) publicly upon publication of the article if not before.

@@

Name your files with evocative and descriptive names. Your
collaborators are less likely to call you at midnight asking for help
if your files are named ``inequality-and-protest-analyses.R'' than if your files are
called ``temp9'' or ``supercalifragilisticexpialidocious'' (Note the
use of the extension .R to tell us that the file contains R
commands. Use extensions like this as a standard practice to help you
and your computer get along.)

@@

I have experimented with three systems so far: (1) for one paper we simply included a Sweave document and data files into a compressed archive \cite{bowers2005dataverse}; (2) for another more computing intensive paper we assembled a set of files which enabled reproduction of our results using the ``make'' system \citep{bowers2008dataverse}; and (3) recently I have tried the ``compendium'' approach \citep{gentleman2005reproducible,gentleman2007statistical} which embeds an academic paper with the R package system \cite{bowers2011dataverse}. The benefit of this last system is that one is not required to have access to unix: the compendium is downloadable from within R using \Verb+install.packages()+ and it viewable using the \Verb+vignette()+ function.

@@

We all always collaborate. Many of us collaborate with groups of people at one moment in time as we race against a deadline. All of us collaborate with ourselves over time. An idealized version of self-collaboration goes something like this: We begin a project with a few notes. We write a conference paper. We revise this paper and send it to a journal. The journal asks for revisions. We revise and resubmit. Our paper is accepted. We revise a bit more. The paper is published. Later another scholar reacts to our paper. We write another paper based on said reactions (draft, conference paper, submission, revision, acceptance, revision). These processes occur over years of time. So, I collaborate with a self 5 or 10 or even more years hence when I write a paper. [What is a reasonable time-span for which to plan for self-collaboration on a single idea? Ask your advisers how long it took them from idea to dissertation to book (articles).]

The time-frames over which collaboration are required --- whether
among a group of people working together or within a single scholar's
productive life --- are much longer than any given version of any
given software will easily exist. Plain text is the exception. Thus,
even as we extol version control systems, one must have a way to
ensure future access to them in a form that will still be around when
the cockroaches take over political science departments.\footnote{
  (and thereby finally make Ctrl-a Ctrl-x Esc-x Ctrl-c a reasonable
  way to instruct Emacs to do something
  (\url{http://kieran.healy.usesthis.com/} )}

Now, imagine that the UN never get's wind of your work, or even that your article does not spawn debate. Why then would you spend time to communicate with your future self and others? My own answer to this question is because I want my work to be credible and useful.  What I report in my data analyses should have two main characteristics: (1) the findings of the work should not be a matter of opinion and (2) other people could reproduce the findings. That is, what the work represents is not a matter of opinion but is a shared experience --- and an experience shared without respect to the identities of others (although requiring some shared technical training and research resources). Such work should make us change how we act --- or at least, it ought to stand on stronger epistemological ground than other claims about experiences which ought to be shared or shareable.

Assume we want others to believe us when we say something. More narrowly, assume we want other people to believe us when we say something about data: ``data'' here can be words, numbers, musical notes, images, ideas, etc \ldots The point is that we are making some claims about patterns that we see in some collection of stuff. Now, it might be easy to convince other people that ``this collection of stuff is different from this collection of stuff'' if those people were looking over our shoulders the whole time that we made decisions about collecting the stuff and analyzed it [where ``analyze'' means something like ``break it up into understandable parts'']. Unfortunately, we can't assume that people are willing to shadow a researcher throughout her career. Rather, we do our work alone or in small groups and want to convince other distant people about our analyses.

In the end, following these practices allows your analyses of your collections to be credible --- and, in the case of criticism, you can always tell antagonists to go redo your analysis if they really don't believe you.  Let me end with a pithy statement in favor of the idea that collaboration is at the heart of our endeavor:
